---
title: 'Workshop 06: More Plots & Normal'
author: 'Wesley Burr'
date: "11/10/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Making Plots for Normal Distributions

Making pictures for the normal problems can help a lot when problem solving. They're also examples of **line graphs** (scatterplots with only one observation per $x$ value), which a few people have been asking about examples for. So let's make one.

Recall from the recent lectures that we have continuous functions which describe probabilities for **continuous random variables**. When working with the Binomial, we used **dbinom()** and **pbinom()**. We're going to do that now, but for the **normal** distribution. The R developers were really careful, and kept most of the syntax the same. The two functions are:

* **pnorm()**: corresponds to P[X $\leq$ q]
* **dnorm()**: the density function - like the P[X = x] from the binomial, but continuous

Let's start by creating a simple plot of the normal density. To make any plot, we need an $x$ and a $y$. All simple
line plots are $y$ versus $x$. What should the $x$ be for this plot ... ?

```{r}
x <- seq(from = -3.5, to = 3.5, by = 0.01)
plot(x = x, y = dnorm(x), type = "l", 
     xlab = "z", ylab = "Normal Density")
```

We can add vertical lines to this:

```{r}
plot(x = x, y = dnorm(x), type = "l",
     xlab = "z", ylab = "Normal Density")
abline(h = 0)
abline(v = 1.1, col = "red")
```

The **abline()** command puts **h** (horizontal) or **v** (vertical) lines onto plots: that's it. We can add more than one line at once by using a vector:

```{r}
plot(x = x, y = dnorm(x), type = "l",
     xlab = "z", ylab = "Normal Density")
abline(h = 0)
abline(v = c(-0.5, 1.1), col = "red")
```

## Solving Problems using **dnorm()** and **pnorm()**

Let's solve some problems involving the standard normal. To do this, **dnorm()** is useful for only the "draw a picture" part of the problem. We have to use **pnorm()**.

### Example 1

If the grades on the SAT are normally distributed, with mean 1500 and standard deviation 275, what is the probability that a randomly selected student will get a score above 1900?

**Solution**: this is asking P[X $\geq$ 1900] for an $X$ which has $\mu=1500$ and $\sigma=275$. We can solve this using the **pnorm()** function. First, let's draw a picture for intuition. 

```{r}
x <- seq(from = 800, to = 2400, by = 0.01)
plot(x = x, y = dnorm(x, mean = 1500, sd = 275), type = "l", 
     xlab = "", ylab = "Normal Density")
axis(side = 1, line = 3, at = seq(700, 2400, 275), 
     labels = round((seq(700, 2400, 275) - 1500) / 275), 4)
abline(h = 0)
abline(v = 1900, col = "red")
```

How big do we think this is? 

Now, solve it using **pnorm()**:
```{r}
pnorm(q = 1900, mean = 1500, sd = 275, lower.tail = FALSE)
```

### Example 2

Suppose that student grades in MATH 1051 are normally distributed with mean 76% and standard deviation 6%. What is the probability that a randomly selected student from the class has a final grade lower than a B (that is, below 70%)?

**Solution**: this is asking P[X < 70] for an $X$ with $\mu = 76$ and $\sigma = 6$. Start with a picture:

```{r}
z <- seq(from = 50.0, to = 100.0, by = 0.1)
plot(x = z, y = dnorm(z, mean = 76, sd = 6), type = "l", 
     xlab = "z", ylab = "Normal Density")
abline(h = 0)
abline(v = c(70), col = "red")
```

This area looks bigger than the last one. Let's guess 0.25. Now, how do we compute it? 

```{r}
pnorm(q = 70, mean = 76, sd = 6, lower.tail = TRUE)
```

So the probability is 0.1586, so the chance of a random student getting less than a B in 1051H, if the grades are actually distributed normally, is 15.9%.

### Challenge Problems

1. A radar unit is used to measure speeds of cars on a motorway. The speeds are normally distributed with a mean of 90 km/hr and a standard deviation of 10 km/hr. What is the probability that a car picked at random is travelling at more than 100 km/hr?
2. The length of similar components produced by a company are approximated by a normal distribution model with a mean of 5 cm and a standard deviation of 0.02 cm. If a component is chosen at random 
    - What is the probability that the length of this component is between 4.98 and 5.02 cm?
    - What is the probability that the length of this component is between 4.96 and 5.04 cm?
3. The time taken to assemble a car in a certain plant is a random variable having a normal distribution of 20 hours and a standard deviation of 2 hours. What is the probability that a car can be assembled at this plant in a period of time 
    - Less than 19.5 hours? 
    - Between 20 and 22 hours? 

## Inverse Probability Problems

The problems we just looked at are **forward** probability problems: we're given a situation, and asked to find the probability. What happens if we have a **backward** probability problem: given a probability, and asked what situation it came from? These are called **inverse** problems.

### Solving an Inverse Problem

The **qnorm()** function does the inverse of the **pnorm()** function. Check out the help system for them now!

```{r}
?qnorm
```

From this, we see that the syntax for qnorm() is very similar to that of pnorm().

```{r, eval = FALSE}
qnorm(p, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
```

So we specify a **p**, the mean and the sd, and then specify which tail we want, and it inverts the problem to get our original $X$ that we're interested in.

### Example 3

The annual salaries of employees in a large company are approximately normally distributed with a mean of \$50,000 and a standard deviation of \$20,000. If the probability of a randomly selected employee making more than a certain amount $X$ per year is 0.21, what is the dollar figure?

**Solution**:
This is saying $P[X $\geq$ x_0] = 0.21$, and we're trying to solve for $x_0$.

```{r}
qnorm(p = 0.21, mean = 50000, sd = 20000, lower.tail = FALSE)
```

Notice how we set **lower.tail = FALSE**, because this is a $\geq$ (greater-than or equal to) problem.

So our solution is \$66,128.42 - so 21% of the company makes more than that amount.

### Example 4

If I tell you that the midterm exam for MATH 1051H had normally distributed scores, with mean 62% and standard deviation 9%, and then
ask you to find the two scores which contain the middle approximately 90% of the class, what are they?

**Solution**:
This is a more complicated problem. Let's slow down, draw a picture, and think a moment.

```{r}
x <- seq(35, 100, 0.01)
plot(x, dnorm(x, mean = 62, sd = 9), type = "l", xlab = "Grade on Midterm", ylab = "Density")
abline(h = 0)
abline(v = c(62-20, 62+20), col = "red")
```

Now, I just made up those numbers (42 and 82) to get lines which cover "about" 90% of the curve. This is to help us think through the problem. We know that the 90% is supposed to be the "middle", which means that since the normal curve is symmetric,
we should have 5% area to the left of the left red line, and 5% area to the right of the right red line, and 90% of the area between them. Write this as a probability statement:

$$
P[X_1 \leq X \leq X_2] = 0.90
$$
and then think about it in terms of $Z$-scores:
$$
Z_1 = \frac{X_1 - 62}{9} \; \qquad \; Z_2 = \frac{X_2 - 62}{9}
$$

But from the symmetry, we know that $X_1$ is the same distance to the left of the mean, $\mu = 62$, as $X_2$ is to the right. So that means the $Z$ scores, which are centered on 0, need to be the same distance as well. In other words,

$$
P[X_1 \leq X \leq X_2] = 0.90 = P[-Z_0 \leq Z \leq Z_0].
$$
And this is the form we need to solve this problem! Because it says that the area to the left of $-Z_0$ is 0.05, just like the area to the right of $+Z_0$.

```{r}
qnorm(p = 0.05, mean = 62, sd = 9, lower.tail = TRUE)
qnorm(p = 0.05, mean = 62, sd = 9, lower.tail = FALSE)
```

So these results tell us that 90% of the class (approximately!) had grades on the midterm of between 47.2% and 76.8%. 

**These are the hardest problems in the first half of the course, so make sure you think about them carefully! They will absolutely be on your midterm (and your final).**

### Challenge Problems

1. Calculate the value of $k$ for which $P[0 < Z < k] = 0.4370$.
2. A manufacturing process produces bolts with mean weight 75g and standard deviation 4g.  10% of the bolts are rejected because they are too heavy. Calculate this weight to the nearest g.
3. Given a normal distribution of values for which the mean is 70 and the standard deviation is 4.5.  Find:
   - the probability that a value is between 65 and 80, inclusive.
   - the 90th percentile for this distribution.

